---
title: "Spatial Processing Benchmarks in R"
author: "Krzysztof Dyba"
output:
  html_document:
    toc: yes
    toc_float: true
---

```{r install-packages, eval=FALSE}
# install all required packages
pkgs = c("bench", "microbenchmark", "sf", "stars", "terra")
install.packages(pkgs)
```

# Introduction

Benchmarking code in R can be done in several ways -- using a built-in function or using more advanced functions from dedicated packages.
The simplest way is the `system.time()` function.
First, we should read the documentation to see how this function works and what it returns.
To view the documentation, use the question mark "?" before the function name.

```{r eval=FALSE}
?system.time
```

We will test it on an example.

```{r}
# generate numbers from normal distribution
# (n = 10,000,000)
t = system.time(rnorm(1e7))
t
```

We are most interested in the third value returned `"elapsed"`, which is the actual execution time of the tested function in seconds (lower value is better/faster).
Of course, we can refer to this value directly using the index in square brackets.

```{r collapse=TRUE}
t[[3]]
t[["elapsed"]]
```

Now let's check out the more advanced functions from the [{bench}](https://bench.r-lib.org/) and [{microbenchmark}](https://cran.r-project.org/web/packages/microbenchmark/index.html) packages on other data.
For the next examples, we can generate a data frame with 3 columns.

```{r}
df = data.frame(x = rnorm(1e5), y = rnorm(1e5), z = rnorm(1e5))
str(df)
```

Benchmarking in the `{bench}` package can be done using the `mark()` function.
It is quite intuitive to call the function explicitly from the package like this: `bench::mark()` (without having to load the entire package).

As a test, we will see how fast can be calculated the sum for each row of a data frame using two different functions: `rowSums()` and `apply()`.

```{r warning=FALSE}
t = bench::mark(
  iterations = 10, check = TRUE,
  rowSums = rowSums(df),
  apply = apply(df, MARGIN = 1, FUN = sum) # 1 indicates rows
)
t[, 1:5]
```

In the `{microbenchmark}` package, we have an analogous `microbenchmark()` function that works similarly, but returns slightly different parameters.

```{r}
library(microbenchmark)
t = microbenchmark(
  times = 10, check = "identical",
  rowSums = rowSums(df),
  apply = apply(df, 1, sum)
)
t
```

We can also easily prepare a visualization (in this case a boxplot).
Moreover, there is `autoplot()` function in the `{ggplot2}` package with different types of plots.

```{r}
boxplot(t)
```

As you can see, both functions allow for timings aggregation (mean, median, etc.) from multiple iterations and results validation, which is a significant improvement over `system.time()`.

This section does not exhaust the topic.
A more complex approach to benchmarks is code profiling, that is, testing the performance of large blocks of code.
This allows us to find out which function is bottleneck.
In R, this can be done using the `Rprof()` function or [{profvis}](https://rstudio.github.io/profvis/) package.
However, for internal C/C++ code, this is not possible (this is true for most packages that are based on GDAL and GEOS).

For more information, see the books:

  - [Advanced R](https://adv-r.hadley.nz/) by Hadley Wickham (chapter [23 Measuring performance](https://adv-r.hadley.nz/perf-measure.html))
  - [Efficient R programming](https://csgillespie.github.io/efficientR/) by Colin Gillespie and Robin Lovelace (chapter [1.6 Benchmarking and profiling](https://csgillespie.github.io/efficientR/introduction.html#benchmarking-and-profiling))

**Exercise** 

Compare computing the mean value of the vector using `mean()` and `sum() / length ()` functions.

# Part I: Raster Data

## Data Source

We will use [Digital Surface Model](https://en.wikipedia.org/wiki/Digital_elevation_model) (DSM) from satellite [Shuttle Radar Topography Mission](https://www2.jpl.nasa.gov/srtm/mission.htm) (SRTM).
You can download the data in 5° x 5° tiles from this website: https://srtm.csi.cgiar.org/srtmdata/.
You can choose any tile, but we will use a tile with ID [**srtm_39_02**](https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_39_02.zip) as an example.
The data is available as a single layer GeoTIFF with a resolution of approximately 90 m (6000 x 6000 pixels).
The coordinate reference system (CRS) of the raster is WGS84 (EPSG:4326), and the datatype is `int16`.

Please be aware that this dataset is quite small and consists of only one layer, so the results are not entirely representative (especially for large-scale applications).

## Benchmarks

In the first step, let's load all the necessary spatial packages.

```{r message=FALSE}
library(sf)
library(stars)
library(terra)
sf_use_s2(FALSE) # use planar GEOS engine by default
```

Next, let's create an empty folder into which we will download the raster.

```{r}
url = "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_39_02.zip"

if (!dir.exists("data")) {
  dir.create("data")
  download.file(url, "data/srtm.zip")
  unzip("data/srtm.zip", exdir = "data") # ~70 MB tif
}

# list files after extracting the archive
list.files("data")
# or save path to raster as variable
f = list.files("data", pattern =  "\\.tif$", full.names = TRUE)
```

In the following sections, we will define the tasks in which we will test our packages.

### Data Loading

Let's start with the simplest task - loading raster data into R.
In the `{stars}` package, the `read_stars()` function is used for this, and in the `{terra}` package, the `rast()` function.

```{r}
x = read_stars("data/srtm_39_02.tif", proxy = FALSE)
x
```

```{r message=FALSE}
plot(x)
```

```{r}
y = rast("data/srtm_39_02.tif")
y
```

```{r}
plot(y)
```

Before we go any further, we need to note one very important point.
In the `rast()` documentation, we can read:

```
When a SpatRaster is created from a file, it does not load the cell (pixel) values into memory (RAM). It only reads the parameters that describe the geometry of the SpatRaster, such as the number of rows and columns and the coordinate reference system. The actual values will be read when needed.
```

So, in fact, when using the `rast()` function, we do not load the raster into memory, but only the metadata describing it.
This is of great importance, because if we do not load it immediately, all the functions tested, will include overhead caused by loading the raster into memory.
To check if the raster has been loaded into memory by `{terra}`, we can use the `inMemory()` function, which will return a logical value.

```{r}
# is the raster in memory?
inMemory(y)
```

As we noted earlier, the raster is not loaded into memory.
To actually load the raster into memory, we need to use the `set.values()` function.

```{r}
set.values(y)
inMemory(y)
```

The situation is similar in the `{stars}` package, but in this case the argument in the `read_stars()` function takes logical values that define whether the file will be directly loaded into memory or not.
See the documentation excerpt below:

```
?read_stars
proxy: logical; if TRUE, an object of class stars_proxy is read which contains array metadata only; if FALSE the full array data is read in memory. (...)
```

After this explanation, we can move on to the benchmark.
The argument `check` in `{bench}` must be obligatorily set to `FALSE`, because the tested packages return different data structures.

```{r}
t = bench::mark(
  iterations = 5, check = FALSE, memory = FALSE,
  stars = read_stars("data/srtm_39_02.tif", proxy = FALSE),
  terra = set.values(rast("data/srtm_39_02.tif"))
)
t[, 1:4]
```

To be fair, we must also note that `{stars}` always reads data as double, and `{terra}` returns integer if the input raster was integer.
This affects the speed of the operation and the amount of RAM memory used.

```{r}
# {stars}
typeof(x[[1]]) # select first attribute `srtm_39_02`
# {terra}
is.int(y) # has raster integer values?
```

### Cropping

In the next task, we will test the raster cropping to the defined range.
We loaded the packages and data in the previous section.
First, let's check the extent of our raster (note that the coordinates in the WGS84 system are in degrees).

```{r}
st_bbox(x) # {stars}
ext(y) # {terra}
```

Next, let's define a new smaller extent to which we will crop the raster.
Note that objects of class `bbox` (`{stars}`) and `ext` (`{terra}`) have different coordinate orders.

```{r}
x_ext = st_bbox(c(xmin = 11, ymin = 51, xmax = 14, ymax = 53),
             crs = st_crs(4326))
y_ext = ext(11, 14, 51, 53)
```

Visualize the new extent:

```{r message=FALSE}
plot(x, axes = TRUE, main = NULL, reset = FALSE)
# bbox object must be converted to sfc geometry
plot(st_as_sfc(x_ext), add = TRUE, border = "red", lwd = 2)
```

Use the `st_crop()` function to crop the raster in `{stars}`, and the `crop()` function in `{terra}`.

```{r}
x_crop = st_crop(x, x_ext)
st_bbox(x_crop)

y_crop = crop(y, y_ext)
ext(y_crop)
```

Everything works fine, so we can move on to the benchmark.
However, earlier we can remove unnecessary objects from the environment using the `rm()` function so that they don't take up space in memory.

```{r warning=FALSE}
rm(x_crop, y_crop)

t = bench::mark(
  iterations = 5, check = FALSE, memory = FALSE,
  stars = st_crop(x, x_ext),
  terra = crop(y, y_ext)
)
t[, 1:4]
```

### Downsampling

The last task will be to reduce the spatial resolution of the raster from higher to smaller (i.e. less detail).
Let's recall the resolution of our input raster.

```{r}
# values in degrees
sapply(st_dimensions(x), "[[", "delta") # {stars}
res(y) # {terra}
```

Next, let's use the `st_warp()` from `{stars}`, and `resample()` from `{terra}`, but before we can do that, we need to prepare the rasters with the target geometry.

```{r warning=FALSE}
x_dest = st_as_stars(st_bbox(x), dx = 0.01, dy = 0.01, values = 0L)
x_small = st_warp(x, x_dest, method = "average", use_gdal = TRUE)
x_small
```

```{r}
y_dest = rast(extent = ext(y), resolution = c(0.01, 0.01), crs = "epsg:4326")
y_small = resample(y, y_dest, method = "average")
y_small
```

Now let's perform the benchmark.

```{r warning=FALSE}
rm(x_small, y_small)

t = bench::mark(
  iterations = 5, check = FALSE, memory = FALSE,
  stars = st_warp(x, x_dest, method = "average", use_gdal = TRUE),
  terra = resample(y, y_dest, method = "average")
)
t[, 1:4]
```

There are other ways to reduce spatial resolution using aggregation.
In `{terra}` you can use the `aggregate()` function which takes the aggregation factor.
In `{stars}` you can use the same function (in fact, this is the equivalent of `terra::extract()`), but you must specify a polygon as an argument -- `st_make_grid()` function can be useful for this.

### Exercise

Prepare a task in which you check the performance of pixel classification using the height condition.
For example, all pixels above the threshold of 200 m will become 1, and all pixels below 0.
This task can be done in many ways, below are some ideas:

```{r eval = FALSE}
# {stars}
## method I
xx = x # duplicate object
xx[xx < 200] = 0
xx[xx >= 200] = 1
typeof(xx[[1]])

## method II (returns factor type, not numeric)
xx = x
xx = cut(xx, breaks = c(-50, 200, 1500), labels = c("0", "1"))
typeof(xx[[1]])

plot(xx, col = c("blue", "red"))
table(xx[[1]]) # frequency table
```

```{r eval = FALSE}
# {terra}
## method I
yy = ifel(y >= 200, 1, 0)

## method II
mat = matrix(c(-50, 200, 0,
               200, 1500, 1),
             nrow = 2, byrow = TRUE)
yy = classify(y, mat)

plot(yy, col = c("blue", "red"))
table(values(yy))
```

# Part II: Vector Data

Attention! If you are lacking RAM after the first part, then you can turn on this notebook directly from the second part (it is independent).

```{r message=FALSE}
library(sf)
library(stars)
library(terra)
sf_use_s2(FALSE) # use planar GEOS engine by default
```

Before starting, let's remove all unnecessary objects from the environment.

```{r}
rm(list = ls())
```

## Data Source

In this section, we will use synthetic (i.e. generated by us) data as input.
As an example, we can use point data.
However, if you would like to use real data, [OpenStreetMap](https://www.openstreetmap.org/) is a great resource.
This data can be obtained using the [{osmdata}](https://cran.r-project.org/web/packages/osmdata/index.html) package.

## Benchmarks
